Parallel Simulation of Adaptive Random Boolean Networks

### 2016-07-14 Paper/paper.tex
Add N_rew and mu_max to Experimental Study

	\paragraph{Algorithm 1 (Liu-Bassler)} This is the algorithm that was used in the original paper introducing Adaptive RBNs \cite{mlb}.  It is based on a set of checkpoints $\mathbf{T} = \{T_0, T_1, T_2, ... T_n\}$. If at some time point $t$ current state $\mathbf{S}(t)$ equals to the state at the checkpoint $\mathbf{S}(T_i)$, an attractor is considered to be found, and some nodes get rewired according to the ADRR presented above. If the program reaches the next checkpoint (except the last one), and $\mathbf{S}(t) \neq \mathbf{S}(T_i)$ $ \forall t \in [T_i, T_{i+1})$, all the information gathered between the checkpoints is discarded as it most probably corresponds to a transient period rather than the attractor, and $T_{i+1}$ becomes the new checkpoint. The longest attractor that can be found using this algorithm equals to the maximum distance between the checkpoints. However, if the distance between the checkpoints is too large, more state updates then necessary are performed. Thus, the values of the checkpoints are crucial to the overall algorithm performance and correctness.

	\paragraph{Algorithm 2 (Knuth)} (The description of Knuth algorithm)

@@

	\begin{equation}
	\label{eq:connectivity}
		K(t) = N^{-1} \sum_{i=1}^{N} {|\mathbf{N}_i(t)|}
	\end{equation}

@@

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/speedup}
			\caption{Execution time of the GPU algorithm compared to the serial one}
			\label{fig:speedup}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/n_rew}
			\caption{Evolution of mean connectivity for different number of nodes rewired each epoch; $N=1000$}
		\end{minipage}
	\end{figure}

	\paragraph{Rewiring more than one node per epoch.}

	\paragraph{Limiting the maximum attractor length.}

	\begin{table}[ht!]
		\centering
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	&	750\\ \hline
			$5 \cdot 10^4$	&	99.77	&	88.19	&	71.19	&	45.48	&	29.62\\
			$2.5 \cdot 10^5$	&	99.99	&	95.92	&	81.24	&	56.10	&	35.96\\
			$5 \cdot 10^5$	&	100.00	&	97.43	&	86.58	&	59.36	&	38.73\\
			$2.5 \cdot 10^6$	&	100.00	&	99.14	&	92.99	&	68.74	&	46.46\\
			$5 \cdot 10^6$	&	100.00	&	99.53	&	94.40	&	73.00	&	49.22\\
			$1 \cdot 10^7$	&	100.00	&	99.75	&	96.08	&	77.73	&	53.81\\
		\end{tabular}
		\caption{Percentage $R$ of attractors found during the evolution process with attractor length limited by $\mu_{max}$}
	\end{table}
	\begin{table}[ht!]
		\centering
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	& 750\\
			\hline
			$5 \cdot 10^4$	&	2.64	&	2.59	&	2.58	&	2.52	&	2.47\\
			$2.5 \cdot 10^5$	&	2.66	&	2.60	&	2.56	&	2.51	&	2.48\\
			$5 \cdot 10^5$	&	2.68	&	2.59	&	2.55	&	2.51	&	2.48\\
			$2.5 \cdot 10^6$	&	2.69	&	2.60	&	2.55	&	2.51	&	2.48\\
			$5 \cdot 10^6$	&	2.69	&	2.59	&	2.55	&	2.52	&	2.48\\
			$1 \cdot 10^7$	&	2.68	&	2.62	&	2.55	&	2.51	&	2.48\\
		\end{tabular}
		\caption{Mean steady state connectivity $K_{ss}$ obtained as a result of evolution process with attractor length limited by $\mu_{max}$}
	\end{table}

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/evolution}
			\caption{Evolution of mean network connectivity}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\textwidth]{plots/attractors}
			\caption{Empirical distribution of attractor lengths}
		\end{minipage}
	\end{figure}
	%------------------------------------------------------------------------------
	\clearpage
	\section{Conclusion}
	The enormous size of RBN ensemble makes them appropriate to model various complex systems but also raises significant challenges for their numerical investigation. Even a problem of search for a single attractor of RBN is computationally hard for a networks larger than several dozens of nodes because of an exponential growth of a number of iterations with an increase of a network size.

>>>>

	\paragraph{Liu-Bassler algorithm.} This is the algorithm that was used in the original paper introducing Adaptive RBNs \cite{mlb}.  It is based on a set of checkpoints $\mathbf{T} = \{T_0, T_1, T_2, ... T_n\}$. If at some time point $t$ current state $\mathbf{S}(t)$ equals to the state at the checkpoint $\mathbf{S}(T_i)$, an attractor is considered to be found, and some nodes get rewired according to the ADRR presented above. If the program reaches the next checkpoint (except the last one), and $\mathbf{S}(t) \neq \mathbf{S}(T_i)$ $ \forall t \in [T_i, T_{i+1})$, all the information gathered between the checkpoints is discarded as it most probably corresponds to a transient period rather than the attractor, and $T_{i+1}$ becomes the new checkpoint. The longest attractor that can be found using this algorithm equals to the maximum distance between the checkpoints. However, if the distance between the checkpoints is too large, more state updates then necessary are performed. Thus, the values of the checkpoints are crucial to the overall algorithm performance and correctness.

	\paragraph{Knuth algorithm.} (The description of Knuth algorithm)

@@

	\begin{equation}
	\label{eq:connectivity}
		K = N^{-1} \sum_{i=1}^{N} {|\mathbf{N}_i|}
	\end{equation}

@@

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/speedup}
			\caption{Execution time of the GPU algorithm compared to the sequential one}
			\label{fig:speedup}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/n_rew}
			\caption{Evolution of mean connectivity for different $N_{rew}$; $N=1000$}
			\label{fig:nrew}
		\end{minipage}
	\end{figure}

	\paragraph{Rewiring more than one node per epoch.} In order to determine how $N_{rew}$ influences the rate of convergence to a dynamical equilibrium, we performed $1000$ epochs of evolution with $N=1000$ and $N_{rew}=[1, 5]$. The results of these simulations are shown in Figure \ref{fig:nrew}. With increased value of $N_{rew}$ mean connectivity grows faster and the system reaches steady state at approximately 500th epoch for $N_{rew}=5$. We limited the attractor length to $\mu_{max}=10^7$ for these experiments.

	\paragraph{Limiting the maximum attractor length.} Introducing a limit on attractor length leads to a situation when an attractor can not be found. In these cases $N_{rew}$ nodes are rewired according to their activity during the last $\mu_max$ state updates. These states can belong either to a transient period or an attractor. However, the results of our experiments, presented in Tables \ref{table:percentage_of_attractors}, \ref{table:kss}, indicate that low percentage of attractors found does not influence the steady state connectivity obtained as a result of the evolution process.

	\begin{table}[ht!]
		\centering
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	&	750\\ \hline
			$5 \cdot 10^4$	&	99.77	&	88.19	&	71.19	&	45.48	&	29.62\\
			$2.5 \cdot 10^5$	&	99.99	&	95.92	&	81.24	&	56.10	&	35.96\\
			$5 \cdot 10^5$	&	100.00	&	97.43	&	86.58	&	59.36	&	38.73\\
			$2.5 \cdot 10^6$	&	100.00	&	99.14	&	92.99	&	68.74	&	46.46\\
			$5 \cdot 10^6$	&	100.00	&	99.53	&	94.40	&	73.00	&	49.22\\
			$1 \cdot 10^7$	&	100.00	&	99.75	&	96.08	&	77.73	&	53.81\\
		\end{tabular}
		\caption{$R$ (percentage of attractors found) for different $N$ and $\mu_{max}$}
		\label{table:percentage_of_attractors}
	\end{table}
	\begin{table}[ht!]
		\centering
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	& 750\\
			\hline
			$5 \cdot 10^4$	&	2.64	&	2.59	&	2.58	&	2.52	&	2.47\\
			$2.5 \cdot 10^5$	&	2.66	&	2.60	&	2.56	&	2.51	&	2.48\\
			$5 \cdot 10^5$	&	2.68	&	2.59	&	2.55	&	2.51	&	2.48\\
			$2.5 \cdot 10^6$	&	2.69	&	2.60	&	2.55	&	2.51	&	2.48\\
			$5 \cdot 10^6$	&	2.69	&	2.59	&	2.55	&	2.52	&	2.48\\
			$1 \cdot 10^7$	&	2.68	&	2.62	&	2.55	&	2.51	&	2.48\\
		\end{tabular}
		\caption{$K_{ss}$ for different $N$ and $\mu_{max}$}
		\label{table:kss}
	\end{table}

	\paragraph{Limiting the maximum attractor length.}
	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/evolution}
			\caption{Evolution of mean network connectivity}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\textwidth]{plots/attractors}
			\caption{Empirical distribution of attractor lengths}
		\end{minipage}
	\end{figure}
	%------------------------------------------------------------------------------
	\clearpage
	\section{Conclusion}
	The enormous size of RBN ensemble makes them appropriate to model various complex systems but also raises significant challenges for their numerical investigation. Even a problem of search for a single attractor of RBN is computationally hard for a networks larger than several dozens of nodes because of an exponential growth of a number of iterations with an increase of a network size.
